Learning Causally-Generated Stationary Time Series

### 2017-02-23 doc/paper.tex
Minor edits, typo's, consistency in style

One of the major goals of statistical inference is to develop models of dynamical phenomena that can be used for prediction, extrapolation, and system identification. Two key characteristics of the physical systems underlying natural and manmade dynamical phenomena are that they are \textit{causal} and \textit{spectrally complex}. Causal systems, where at any point in time the output of the system can only depend on past values of the input, are the only type which are physically realisable. Spectrally complex systems, which have rich power-spectral densities, arise because many physical systems have numerous resonances spanning many time scales. When developing statistical models for such phenomena this prior knowledge---causality and spectral richness---should be leveraged in order to exclude any unrealisable system from the model prior whilst at the same time allowing the model to have the capacity to capture varied spectral content that may be only slowly revealed as more data are seen. The goal of this paper is to develop Gaussian Process models together with associated inference and learning schemes that serve this purpose.

Gaussian Processes are a widely-used model for a stationary time series that place a prior distribution over the latent function underlying the time series $f:\R \to \R$ by assuming that any finite collection of function values $f(t_1),\ldots,f(t_n)$ is multivariate-Gaussian distributed. The key modelling decision in using a Gaussian process is the choice of covariance  $k_f(t,t')$ between any two function values $f(t)$ and $f(t')$---$k_f$ is also called the \textit{kernel} of $f$. This choice of kernel\footnote{The mean of the function is assumed to be zero without loss of generality. \cite{Rasmussen:2006:Gaussian_Processes}} encodes prior information about the function $f$. Consequently, a large research effort has been devoted toward developing flexible kernels \cite{Duvenaud:2014:Automatic_Construction,Wilson:2013:Spectral_Mixture,Tobar:2015:Learning_Stationary,Tobar:2015:Inter-Domain_Inducing}.

Of particular interest for the current application are GP models that use non-parametric models of the power-spectra of the Gaussian Process (doubly non-parametric models) as they have the capacity to flexibly model signals with arbitrary spectral complexity. One approach employs a Dirichlet Process Mixture Model for the power spectra \cite{Bayesian nonparametric kernel learning, Wilson}, but it is not clear whether it is possible to build causality into such a construction. An alternative approach induces a non-parametric power spectrum by placing a Gaussian Process prior over the system's impulse response function  \citet{Tobar:2015:Learning_Stationary}. Critically, the so-called Gaussian Process Convolution Model (GPCM), does not incorporate a causality constraint (the impulse response in question should be thought of as an acausal spatial response). In this paper we revisit the GPCM and explicitly build in a causal constraint whilst retaining the flexibility of modelling the kernel nonparametrically. In addition to these modelling contributions, we develop enhanced variational inference and learning schemes for the causal and acausal variants of the GPCM -- collapsed variational bounds and structured approximating distributions -- that significantly improve statistical accuracy. These modelling and inferential contributions are demonstrated on synthetic and real world signals.

>>>>

One of the major goals of statistical inference is to develop models of dynamical phenomena that can be used for prediction, extrapolation, and system identification. Two key characteristics of the physical systems underlying natural and manmade dynamical phenomena are that they are \textit{causal} and \textit{spectrally complex}. Causal systems, where at any point in time the output of the system can only depend on past values of the input, are the only type which are physically realisable. Spectrally complex systems, which have rich power spectral densities, arise because many physical systems have numerous resonances spanning many time scales. When developing statistical models for such phenomena this prior knowledge---causality and spectral richness---should be leveraged in order to exclude any unrealisable system from the model prior whilst at the same time allowing the model to have the capacity to capture varied spectral content that may be only slowly revealed as more data are seen. The goal of this paper is to develop Gaussian Process (GP) models together with associated inference and learning schemes that serve this purpose.

A widely-used model for stationary time series are Gaussian Processes, which place a prior distribution over the latent function underlying the time series $f:\R \to \R$ by assuming that any finite collection of function values $f(t_1),\ldots,f(t_n)$ is multivariate-Gaussian distributed. The key modelling decision in using a Gaussian Process is the choice of covariance $k_f(t,t')$ between any two function values $f(t)$ and $f(t')$---$k_f$ is also called the \textit{kernel} of $f$. This choice of kernel\footnote{The mean of the function is assumed to be zero without loss of generality. \cite{Rasmussen:2006:Gaussian_Processes}} encodes prior information about the function $f$. Consequently, a large research effort has been devoted towards developing flexible kernels \cite{Duvenaud:2014:Automatic_Construction,Wilson:2013:Spectral_Mixture,Tobar:2015:Learning_Stationary,Tobar:2015:Inter-Domain_Inducing}.

Of particular interest for the current application are GP models that use nonparametric models of the power spectra of the GPs---doubly non-parametric models---as they have the capacity to flexibly model signals with arbitrary spectral complexity. One approach employs a Dirichlet Process Mixture Model for the power spectra \cite{Oliva:2015:Bayesian_Nonparametric_Kernel-Learning}, but it is not clear whether it is possible to build the causality constraint into such a construction. An alternative approach instead induces a nonparametric power spectrum by placing a Gaussian Process prior over the system's impulse response function \cite{Tobar:2015:Learning_Stationary}. Critically, the so-called Gaussian Process Convolution Model (GPCM), does not incorporate a causality constraint---the impulse response in question should be thought of as an acausal spatial response. In this paper we revisit the GPCM and explicitly build in a causality constraint whilst retaining the flexibility of modelling the kernel nonparametrically. In addition to these modelling contributions, we develop enhanced variational inference and learning schemes for the causal and acausal variants of the GPCM---collapsed variational bounds and structured approximating distributions---that significantly improve statistical accuracy. These modelling and inferential contributions are demonstrated on synthetic and real-world signals.
